---
title: Learn Stan with brms, Part II
subtitle: "bf(y ~ 1 + x, center = FALSE)"
author: A. Solomon Kurz
date: '2025-07-13'
excerpt: "In the first post of this series, we focused on a single-level intercept-only model. In this post we expand the model to include a single continuous predictor. Along the way, we review matrix notation, model matrices, and a new class of distribution functions for Stan."
tags:
  - Bayesian
  - brms
  - gaussian
  - GLM
  - regression
  - rstan
  - stan
draft: false
layout: single
featured: no
bibliography: /Users/solomonkurz/Dropbox/blogdown/content/blog/my_blog.bib
biblio-style: apalike
csl: /Users/solomonkurz/Dropbox/blogdown/content/blog/apa.csl  
link-citations: yes
nocite: |  # This adds references not cited in the prose
  @R-base, @R-tidyverse, @wickhamWelcomeTidyverse2019, @R-ggdist, @R-brms, @burknerBrmsPackageBayesian2017, @burknerAdvancedBayesianMultilevel2018, @R-rstan, @R-posterior, @R-tidybayes
---

## We can predict

In the [first post](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/) of this series, we focused on a single-level intercept-only model. Even in that context, the Stan code used by `brm()` included several sophisticated features. In this post we expand the model to include a single continuous predictor. Along the way, we review matrix notation, model matrices, and a new class of distribution functions for Stan. 

Load the primary **R** packages and the data file.

```{r, eval = F}
# Packages
library(tidyverse)
library(brms)
library(rstan)
library(tidybayes)
library(ggdist)
library(posterior)

# Data
load(file = "data/d.rda")

# Review
glimpse(d)
```

```{r, warning = F, message = F, echo = F}
# Packages
library(tidyverse)
library(brms)
library(rstan)
library(tidybayes)
library(ggdist)
library(posterior)

# Data
load(file = "/Users/solomonkurz/Dropbox/blogdown/content/blog/2025-07-07-learn-stan-with-brms-part-i/data/d.rda")

# Review
glimpse(d)
```

We continue with the height and weight data. For the code underlying the data, review the [appendix](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#appendix) from the first post.

## Focal model

In this post we will fit the univariable model

$$
\begin{align}
\text{weight}_i & \sim \operatorname{Normal}(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 \text{height}_i \\
\beta_0 & \sim \operatorname{Normal}(0, 100) \\
\beta_1 & \sim \operatorname{Normal}(5, 2.5) \\
\sigma &  \sim \operatorname{Exponential}(0.04),
\end{align}
$$

where `weight` values vary across $i$ people, and they are modeled as normally distributed with a conditional mean $\mu_i$ and [residual] standard deviation $\sigma$. The conditional mean is defined by an intercept $\beta_0$ and slope $\beta_1$ for the predictor variable `height`. This time we discuss the priors in reverse order.

The $\sigma$ parameter has changed meaning from a unconditional standard deviation to a conditional or residual standard deviation. Before we used an exponential prior centered at 25 [cf. @mcelreathStatisticalRethinkingBayesian2020], and for simplicity here we do the same. Recall that the mean of the exponential distribution is the reciprocal of its single parameter $\lambda$, and we therefore set $\lambda = 1/25 = 0.04$.

Our new $\beta_1$ parameter is the expected change in `weight` for a one-unit change^[Keep in mind *change* here is only used descriptively. Maybe `height` has a simple causal relation to `weight`, maybe not.] in the predictor `height`. Given a normal prior, the values of a mean of five and standard deviation of 2.5 may seem oddly specific. A few things: My experience as a human living in the world among other humans tells me the $\beta_1$ parameter will be positive, so a good theory-informed priors should down rate negative values. When I did a quick web search on the relation between height and weight, @peterson2016universal popped up which indicated that given the typical adult range for height, a one inch increase predicts about a 5 lb increase in ideal weight. By scaling the prior with 2.5, that puts about 95% of the prior probability between zero and twice that value.

In this model, I think $\beta_0$ is the most challenging parameter for which to set a prior. The intercept, recall, is the expected value when the predictor variable is set to zero. At a first stab, you might insist that a hypothetical person zero inches tall would have to also weight zero pounds. But bear in mind we are fitting a simple linear model, which has no simple way to enforce such a constraint. For that we'd probably want some kind of exciting non-linear model.^[You can find several non-linear strategies for these kinds of data in @mcelreathStatisticalRethinkingBayesian2020.] Also, given that we are using `weight` in inches, and given these data were simulated to resemble typical US adults, none of the `weight` values are remotely close to zero, which means that not only we, but the model itself will have a very difficult time estimating the expected value at zero. So to express this uncertainty, but to still suggest our best guess is the value will be around zero, we use a normal prior centered at zero, with a large scale of 100.

Here's a quick look at the priors.

```{r, fig.width = 8, fig.height = 3.25, warning = FALSE}
c(prior(normal(0, 100), class = b, coef = Intercept),
  prior(normal(5, 2.5), class = b),
  prior(exponential(0.04), class = sigma)) |> 
  parse_dist() |> 

  ggplot(aes(y = prior)) +
  stat_slab(aes(dist = .dist_obj),
            n = 5001, normalize = "xy", p_limits = c(0.0001, 0.9999)) +
  scale_y_discrete(NULL, expand = expansion(mult = 0.1)) +
  xlab("parameter space") +
  coord_cartesian(xlim = c(-325, 325)) +
  theme(panel.grid = element_blank())
```

### Center the predictor

Some of my more experienced readers might wonder why I haven't made things easier on myself by centering the predictor variable. Then we could drop that annoyingly wide $\operatorname{Normal}(0, 100)$ prior for $\beta_0$. Yes, that's a great strategy in many contexts. Mean centering predictor variables is a *big deal* with **brms**, particularly with how you set your priors.^[In more ways than one.] It's such a big deal that it will be the focus of the [next post](https://solomonkurz.netlify.app/blog/2025-07-17-learn-stan-with-brms-part-iii/). But the mean-centering issue and how it relates to **brms** syntax complicates the underlying Stan code in ways I think would be prohibitively confusing at this point. The Stan code is simpler for a model fit with a non-standardized predictor. And plus, it's nice to learn how to fit a Stan model with a non-centered predictor. As an applied researcher, I like it when I have options.

## **brms** workflow

For **brms** users who've read the first post in this series, there should be few surprises here.

### Fit the model with `brm()`

We can fit this model with **brms** like so.

```{r brm2}
brm2 <- brm(
  data = d,
  family = gaussian,
  bf(weight ~ 1 + height, center = FALSE),
  prior = prior(normal(0, 100), class = b, coef = Intercept) +
    prior(normal(5, 2.5), class = b) +
    prior(exponential(0.04), class = sigma),
  cores = 4, seed = 1,
  file = "fits/brm2")
```

The only noteworthy part of the model code is how we wrapped our formula line in the `bf()` function, in which we also set `center = FALSE`. Those details have to do with setting the prior for $\beta_0$ using the current version of `height` that is *not* mean centered. You can execute `?brmsformula` in your console for the technical documentation. We'll have more to say on that whole issue in the [next post](https://solomonkurz.netlify.app/blog/2025-07-17-learn-stan-with-brms-part-iii/). For now, just go with it.

Check the parameter summary.

```{r}
print(brm2)
```

We might use the posterior draws to plot the fitted line against the data. To better understand the role of the intercept $\beta_0$, we will extend the regression line well past the lower limit of the data all the way to zero. For kicks and giggles, we'll also superimpose the OLS regression line in red.

```{r, fig.width = 8, fig.height = 4}
xmax <- max(d$height) |> ceiling()

data.frame(height = 0:xmax) |> 
  add_epred_draws(brm2) |> 
  rename(weight = .epred) |> 
  
  ggplot(aes(x = height, y = weight)) +
  stat_lineribbon(.width = 0.95,
                  color = "steelblue4", fill = "steelblue1", linewidth = 1/2) +
  geom_point(data = d) +
  stat_smooth(data = d,
              formula = 'y ~ x', method = "lm", se = FALSE,
              color = "red") +
  theme(panel.grid = element_blank())
```

In this case, our Bayesian regression line matched nicely with the OLS line.^[I compare the Bayesian line to OLS not because OLS is the benchmark. It isn't. But given how small the data cloud is in this plot, it's hard to judge how well the line is describing the data, and one could easily wonder how strongly the $\beta$ priors are influencing the posterior. Yes, the line describes the data well. No, these particular priors aren't pushing the posterior that far from the likelihood.] But also, hopefully it now makes sense why we used such a weak prior for $\beta_0$. The observed data are long ways from zero, and there's no good way to enforce an exact-zero intercept with a simple linear model.^[I mean, sure, you could do so by dropping the $\beta_0$ term altogether, which would effectively constrain the intercept to zero. But (a) that would do a worse job describing the observed data, and (b) it would no longer be a simple liner model, which is our framework here.]

```{r, fig.width = 8, fig.height = 4, echo = F, eval = F}
xmax <- max(d$height) |> ceiling()

brm2.1 <- brm(
  data = d,
  family = gaussian,
  weight ~ 0 + height,
  prior = prior(normal(5, 2.5), class = b) +
    prior(exponential(0.04), class = sigma),
  cores = 4, seed = 1)

print(brm2.1)

data.frame(height = 0:xmax) |> 
  add_epred_draws(brm2.1) |> 
  rename(weight = .epred) |> 
  
  ggplot(aes(x = height, y = weight)) +
  stat_lineribbon(.width = 0.95,
                  color = "steelblue4", fill = "steelblue1", linewidth = 1/2) +
  geom_point(data = d) +
  stat_smooth(data = d,
              formula = 'y ~ x', method = "lm", se = FALSE,
              color = "red") +
  theme(panel.grid = element_blank())
```

### Extract the Stan code with `stancode()`

As we learned in the first post, we can extract the Stan code underlying our `brm2` model with the `stancode()` function.

```{r}
stancode(brm2)
```

Many of the features in this Stan code mirror those from the Stan code from the [first post](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#extract-the-stan-code-with-stancode). But if you look closely, there are some new features in the `data`, `parameters` and `model` blocks, and now the `transformed data` and `generated quantities` blocks are empty! We will discuss those new features in the sections below.

### Extract the Stan data with `standata()`

Investigate the data **brms** actually passed onto Stan with the `standata()` function.

```{r}
standata(brm2) |> 
  str()
```

This is largely the same as in the [last post](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#extract-the-stan-data-with-standata). In the sections below, we will learn about the mysterious `X` and `K` elements.

## **rstan** workflow

We will fit four **rstan** versions of the `brm2` model. We start simple with the first model `stan2.1`. We then take a whirlwind review of the major Stan-code insights from the last post with `stan2.2`. With `stan2.3` we learn about model matrices and parameter vectors, and with `stan2.4` we meet a new family of functions parameterized for the generalized linear model.

### `stan2.1`, start simple

As a first step, we make a data list with the `compose_data()` function.

```{r}
stan_data <- d |> 
  select(weight, height) |> 
  compose_data()

str(stan_data)
```

The `weight` and `height` columns have been saved as named numeric vectors in the list, and `compose_data()` automatically added the `n` scalar value defining the number of cases in the data. We don't have analogues to the `X` and `K` elements yet, but those will come [later][Data matrices and parameter vectors in Stan].

As in the last post, our first `stan()` model will use a simplified version of Bürkner's Stan code. You might compare this with the first example in the [*Regression Models*](https://mc-stan.org/docs/stan-users-guide/regression.html) chapter of the *Stan User's Guide* [@standevelopmentteamStanUserGuide2024].

```{r}
model_code_2.1 <- '
data {
  int<lower=1> n;    // Total number of observations
  vector[n] weight;  // Response variable
  vector[n] height;  // Predictor variable
}
parameters {
  real beta0;           // Intercept
  real beta1;           // Slope
  real<lower=0> sigma;  // Dispersion parameter
}
model {
  weight ~ normal(beta0 + beta1 * height, sigma);  // Likelihood
  beta0 ~ normal(0, 100);                          // Priors
  beta1 ~ normal(5, 2.5);
  sigma ~ exponential(0.04);
}
'
```

In the `data` block we have added a line for the predictor variable `height`, which is a vector of length `n`. In the `parameters` block we now have an intercept and a slope, which we've defined on separate lines. We did not have to choose the names `beta0` and `beta1`, and you are at liberty to alter them if you're following along on your own computer. Notice we're using the *distribution* syntax for the likelihood and priors in the `model` block. We'll switch that out for more sophisticated syntax in the models to come. In the likelihood line, notice how we're defining the entire linear model right there in the first parameter of `normal()`. There are other ways to do this, and we'll explore those below. But for now, notice how we used the syntax of `beta1 * height`, where we multiply the `beta1` slope with the `height` vector using the `*` operator. This is more explicit than the typical `1 + height` syntax we use in **R** functions like `brm()` or `lm()`. At this point, I'm hoping the syntax we used for the priors will make sense.

We're ready to fit the model with `stan()`.

```{r stan2.1, echo = F}
# save(stan2.1, file = "fits/stan2.1.rda")
load(file = "fits/stan2.1.rda")
```

```{r, eval = F}
stan2.1 <- stan(
  data = stan_data,
  model_code = model_code_2.1,
  cores = 4, seed = 1)
```

The `print()` summary for our `stan()` model is very similar to the output for `brm2` above, but it's not identical. By the end of this post, we will perfectly match the draws from `brm2`.

```{r}
# summary(brm2)
print(stan2.1, probs = c(0.025, 0.975))
```

### `stan2.2` and a quick Stan code review

We'll use this section to review all the Stan syntax insights we learned in the first post, and then add them to the model in quick succession.

#### `lprior` and `target += normal_lpdf`

Our first step is to define the `lprior` vector in the `transformed parameters` block. We also switch to the *log probability density function* syntax for the likelihood and priors, with its exciting looking `+=` operator. In short, the `lprior` is the log of the prior density, and `target` is a reserved term used for the log of the posterior density. For more details on these additions, go back and review [this section](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#stan12-by-adding-a-target) from the first post. 

```default
data {
  int<lower=1> n;
  vector[n] weight;
  vector[n] height;
}
parameters {
  real beta0;
  real beta1;
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;  // Prior contributions to the log posterior
  lprior += normal_lpdf(beta0 | 0, 100);
  lprior += normal_lpdf(beta1 | 5, 2.5);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  target += normal_lpdf(weight | beta0 + beta1 * height, sigma);  // Likelihood
  target += lprior;                                               // Priors
}
```

You can fit a model using this `model_code` as an intermediate step, but I'm going to press forward.

#### `prior_only` and `if`

Next we can add that `if` statement to the `model` block that can handle the `prior_only` option (*Should we sample from the posterior, or from the prior?*). For more details on these additions, go back and review [this section](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#stan16-for-prior_only-and-if) from the first post.

```default
data {
  int<lower=1> n;
  vector[n] weight;
  vector[n] height;
  int prior_only;  // Should the likelihood be ignored?
}
parameters {
  real beta0;
  real beta1;
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(beta0 | 0, 100);
  lprior += normal_lpdf(beta1 | 5, 2.5);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  // Likelihood
  if (!prior_only) {
    target += normal_lpdf(weight | beta0 + beta1 * height, sigma);
  }
  // Priors
  target += lprior;
}
```

For this step, we need to add that `prior_only` scalar to the data list, which we do like so.

```{r}
stan_data <- d |> 
  select(weight, height) |> 
  compose_data(prior_only = as.integer(FALSE))

str(stan_data)
```

You can fit a model using this `model_code` as an intermediate step, but I'm going to press forward, still. We have one more point to review, and this time that review should relieve an unresolved tension from the first post. Get excited!

#### `mu += Intercept`, but better

You might recall that back in the first post ([here](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#stan15-adds-a-b_intercept)), we discussed this `mu += Intercept` line Bürkner smuggled into the middle of his `model` block. It didn't make any sense in the context of an intercept-only model, but we followed along anyway and let the tension build. Take another look at the current state of our likelihood line.

```default
target += normal_lpdf(weight | beta0 + beta1 * height, sigma);
```

The linear function in the model is starting to take up a lot of horizontal space, and you could only imagine how this might spiral out of control if we were to add more predictor terms to the model. So one nice trick is to define a new term `mu`, and then just substitute that term into the `normal_lpdf()` function.

```default
vector[n] mu = rep_vector(0.0, n);
mu += beta0 + beta1 * height;
target += normal_lpdf(weight | mu, sigma);
```

Notice how before we define `mu` with the linear model, we first declare its dimensions in the opening line `vector[n] mu = rep_vector(0.0, n)`. When we use code like this within a `model` block, the posterior will *not* return samples from `mu`. It will just return draws from the `beta0` and `beta1` parameters instead. But if we did want to sample from `mu` as well, we could define it in the `transformed parameters` or `generated quantities` blocks.

For our use case, here's what this looks like in full.

```{r}
model_code_2.2 <- '
data {
  int<lower=1> n;
  vector[n] weight;
  vector[n] height;
  int prior_only;
}
parameters {
  real beta0;
  real beta1;
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(beta0 | 0, 100);
  lprior += normal_lpdf(beta1 | 5, 2.5);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  // Likelihood
  if (!prior_only) {
    // Initialize linear predictor term
    vector[n] mu = rep_vector(0.0, n);
    mu += beta0 + beta1 * height;
    target += normal_lpdf(weight | mu, sigma);
  }
  // Priors
  target += lprior;
}
'
```

Now the line for the likelihood, `target += normal_lpdf(weight | mu, sigma)`, no longer takes up all that horizontal space. And hopefully now you can also make sense of Bürkner's comment: `Initialize linear predictor term`. `mu` is the linear predictor term. Time to fit the model with `stan()`.

```{r stan2.2, echo = F}
# save(stan2.2, file = "fits/stan2.2.rda")
load(file = "fits/stan2.2.rda")
```

```{r, eval = F}
stan2.2 <- stan(
  data = stan_data,
  model_code = model_code_2.2,
  cores = 4, seed = 1)
```

The only major change in the `print()` output is the addition of the `lprior` line. Again, notice how there is no line for `mu`. Because we only defined `mu` in the `model` block, it was just temporary.

```{r}
# summary(brm2)
# print(stan2.1, probs = c(0.025, 0.975))
print(stan2.2, probs = c(0.025, 0.975))
```

### `stan2.3` with `model.matrix()`

It's time to learn some new Stan syntax tricks. Before we do, we'll want to discuss some statistical notation.

#### Compact notation and design matrices

When I express my models in statistical notation, I prefer the style

$$
\mu_i = \beta_0 + \beta_1 x_i
$$

for models with a single predictor $x$. When I have multiple predictors, I just generalize to

$$
\mu_i = \beta_0 + \beta_1 x_{1i} + \dots + \beta_K x_{Ki},
$$

where $k$ is a predictor index and $K$ is the total number of predictor variables. I find this notation style simple and easy to read, and of all the ways to write a linear model, this is the one I suspect will be the most accessible to my substantive colleagues.^[I'm a clinical psychologist, recall. My fellow clinical psychologists generally don't care for equations, but if you're going to force an equation on them, this is the style I'd recommend.] However, this style of equation doesn't work well when $K$ increases beyond the middle single-digit range.

One solution is the notation of 

$$
\mu_i = \alpha + \sum_{k = 1}^K \beta_k x_{ki},
$$

where we now refer to the intercept as $\alpha$, and we use the summation operator $\sum(\cdot)$ as a shorthand for $\beta_1 x_{1i} + \dots + \beta_K x_{Ki}$. We can use an even more compact notation of

$$
\mathbf m = \mathbf{Xb}.
$$

This is matrix notation, for which $\mathbf m$ is a vector of predicted means which is the same length as the sample data (100 rows in our case). The $\mathbf X$ term is a matrix of predictor values, which includes one column for each of the predictor variables *and* an initial column that is a vector of 1's for the intercept. The $\mathbf b$ term is a vector of parameters, which begins with the intercept and ends with the final predictor $x_K$. We could try to write this out more explicitly as something like

$$\begin{bmatrix} \mu_1 \\ \mu_2 \\ \vdots \\ \mu_N \end{bmatrix} = \begin{bmatrix} 1 & x_{11} & \dots & x_{K1} \\ 1 & x_{12} & \dots & x_{K2} \\ \vdots & \vdots & \ddots & \vdots \\ 1 & x_{1N} & \dots & x_{KN} \end{bmatrix} \begin{bmatrix} \beta_0 \\ \beta_1 \\ \vdots \\ \beta_K \end{bmatrix}.$$

#### Data matrices and parameter vectors in Stan

In **R**, we can define a model matrix with the `model.matrix()` function.^[Execute `?model.matrix` for the documentation.] For our model, that can look like so.

```{r}
# Define a model matrix
mm_2.3 <- model.matrix(data = d, object = ~height)

# What?
str(mm_2.3)
head(mm_2.3)
```

The `mm_2.3` object has two vectors: The first vector is a sequence of `1`'s for the intercept, and the second vector has the values for the predictor `height`. In the notation above, these would be

$$\begin{bmatrix} 1 & x_{11} \\ 1 & x_{12} \\ \vdots & \vdots \\ 1 & x_{1N} \end{bmatrix},$$

where $x_1$ is `height`, and $N$ is our scalar value `n`. We can use this `mm_2.3` object to define two new elements with the `compose_data()` function. The first will be `X`, to stand for the entire model matrix `mm_2.3`. The second will be `k`, the number of columns (i.e., "predictors") in the design matrix.

```{r}
stan_data <- d |> 
  select(weight) |> 
  compose_data(k = ncol(mm_2.3),
               X = mm_2.3,
               prior_only = as.integer(FALSE))

str(stan_data)
```

It's important to notice that up until this point we've only been referring to variables like $x_1$ (`height`) as *predictors*. But with this kind of model matrix, we are also including the vector of 1's for the intercept as a "predictor."  Thus in that sense, our model matrix has `k == 2` predictors.

Notice how because we have added our `height` values to the data list within the `X` model matrix, we no longer add them in from the data frame `d`. That is, we now just use `d |> select(weight)` in the opening lines of the code.

We can add this information to the `data` block of our `model_code` like so.^[See also the [*Matrix notation and vectorization*](https://mc-stan.org/docs/stan-users-guide/regression.html#vectorization.section) section in the *Stan User's Guide* [@standevelopmentteamStanUserGuide2024].]

```{r}
model_code_2.3 <- '
data {
  int<lower=1> n;
  int<lower=1> k;  // Number of beta coefficients in the linear model
  vector[n] weight;
  matrix[n, k] X;  // Design matrix for the beta coefficients
  int prior_only;
}
parameters {
  vector[k] b;  // Regression coefficients (beta coefficients)
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(b[1] | 0, 100);
  lprior += normal_lpdf(b[2] | 5, 2.5);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  if (!prior_only) {
    // Linear model defined in matrix algebra notation Xb
    target += normal_lpdf(weight | X * b, sigma);
  }
  target += lprior;
}
'
```

We have several points of interest. Focus on the `data` block, first.

```default
data {
  int<lower=1> n;
  int<lower=1> k;  // Number of beta coefficients in the linear model
  vector[n] weight;
  matrix[n, k] X;  // Design matrix for the beta coefficients
  int prior_only;
}
```

We have declared our `k` scalar in the second line. We no longer declare the `height` values as a vector, like we did in earlier models. But now instead we declare the design matrix `X` as a matrix of `n` rows and `k` columns.

Next focus on the `parameters` block.

```default
parameters {
  vector[k] b;  // Regression coefficients (beta coefficients)
  real<lower=0> sigma;
}
```

We have dropped those two lines that used to declare `beta0` and `beta1`. Now we instead have a vector of `b` coefficients of length `k`, which is the first time we've seen a vector of parameters in this blog series. This is the code instantiation of what we might have described in statistical notation above as

$$\begin{bmatrix} \beta_0 \\ \beta_1 \end{bmatrix}.$$

As there's nothing new for us in the `transformed parameters` block, we finally focus instead on the `model` block.

```default
model {
  if (!prior_only) {
    // Linear model defined in matrix algebra notation Xb
    target += normal_lpdf(weight | X * b, sigma);
  }
  target += lprior;
}
```

Exciting as it was in the last section, we have dropped all that `mu += beta0 + beta1 * height` stuff. Now we instead define the entire linear model with the compact syntax of `X * b`, which you might notice looks a lot like the matrix algebra notation $\mathbf{Xb}$ we just learned about in the equations above. That is, we have specified the linear model

$$\mathbf{Xb} = \begin{bmatrix} 1 & x_{1,1} \\ 1 & x_{1,2} \\ \vdots & \vdots \\ 1 & x_{1,100} \end{bmatrix} \begin{bmatrix} \beta_0 \\ \beta_1 \end{bmatrix}.$$

We're ready to fit the model with `stan()`.

```{r stan2.3, echo = F}
# save(stan2.3, file = "fits/stan2.3.rda")
load(file = "fits/stan2.3.rda")
```

```{r, eval = F}
stan2.3 <- stan(
  data = stan_data,
  model_code = model_code_2.3,
  cores = 4, seed = 1)
```

Check the summary.

```{r}
# summary(brm2)
# print(stan2.1, probs = c(0.025, 0.975))
# print(stan2.2, probs = c(0.025, 0.975))
print(stan2.3, probs = c(0.025, 0.975))
```

Overall, the parameter summaries are very similar to those from `stan2.2`. But notice the names of the first two rows. When we use this approach, all $\beta$ coefficients in the model now take generic names `b[k]`, ranging from $1, \dots, K$. Even though we^[Or at least me, Solomon, anyway.] often refer to the intercept as $\beta_0$, and so on, here the intercept is `b[1]`. To the best of my knowledge, there is no simple way to fix this when using this approach to fitting models with Stan. The first index in a vector is a 1, not a zero. We won't have a good solution to that challenge in this post, but we will present an alternative in the [third post](https://solomonkurz.netlify.app/blog/2025-07-17-learn-stan-with-brms-part-iii/). For now, let's prepare for one last Stan syntax trick.

### `stan2.4` with the `normal_id_glm_lpdf()` function

Here's the final version of the `model_code` for this post.

```{r}
model_code_2.4 <- '
data {
  int<lower=1> n;
  int<lower=1> k;
  vector[n] weight;
  matrix[n, k] X;
  int prior_only;
}
parameters {
  vector[k] b;
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(b[1] | 0, 100);
  lprior += normal_lpdf(b[2] | 5, 2.5);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  if (!prior_only) {
    target += normal_id_glm_lpdf(weight | X, 0, b, sigma);
  }
  target += lprior;
}
'
```

The only new line is for the likelihood in the `model` block. Following the syntax in Bürkner's Stan code, we have abandoned the `normal_lpdf()` function for `normal_id_glm_lpdf()`. Stan provides a family of functions parameterized for the generalized linear model. This family of functions follows the naming convention of `distribution_link_glm_function()`. In this case, we are using the Gaussian distribution (`normal`), the identity link (`id`), and the log probability density function (`lpdf`). For the documentation, go [here](https://mc-stan.org/docs/functions-reference/unbounded_continuous_distributions.html#normal-id-glm) for the relevant section of the *Stan Functions Reference* [@standevelopmentteamStanFunctionsReference2024]. On the left side of the `|` bar, we input the response variable `weight`. The four arguments on the right side of the `|` bar are $x, \alpha, \beta, \sigma$. For the first of those four, we input our model matrix `X`, which includes a constant vector of `1` for the intercept and a vector of the numeric values of the predictor `height`. In our case, we are not using the $\alpha$ argument because we have included the intercept in `X`. In the [next post](https://solomonkurz.netlify.app/blog/2025-07-17-learn-stan-with-brms-part-iii/#stan32-for-new-tricks-and-the-final-model), however, we will see an example of how we can use $\alpha$. But notice we do input our vector of `b` parameters in the $\beta$ argument. Perhaps not surprisingly at this point, we input our `sigma` parameter in the final argument $\sigma$. In the *Stan Functions Reference*, we read this function is designed to provide "a more efficient implementation of linear regression than a manually written regression in terms of a normal distribution and matrix multiplication." 

Let's fit the model and see.

```{r stan2.4, echo = F}
# save(stan2.4, file = "fits/stan2.4.rda")
load(file = "fits/stan2.4.rda")
```

```{r, eval = F}
stan2.4 <- stan(
  data = stan_data,
  model_code = model_code_2.4,
  cores = 4, seed = 1)
```

Check the summary.

```{r}
# summary(brm2)
# print(stan2.1, probs = c(0.025, 0.975))
# print(stan2.2, probs = c(0.025, 0.975))
# print(stan2.3, probs = c(0.025, 0.975))
print(stan2.4, probs = c(0.025, 0.975))
```

Was the `normal_id_glm_lpdf()` method efficient in terms of speed relative to the `normal_lpdf()` method? We can extract the warmup and sample times with the `get_elapsed_time()` function.

```{r}
get_elapsed_time(stan2.3)  # `normal_lpdf()` method
get_elapsed_time(stan2.4)  # `normal_id_glm_lpdf()` method
```

Yes, it turns out the `normal_id_glm_lpdf()` method was indeed more efficient for both stages, and across all four chains.

```{r, echo = F, eval = F}
# No notable difference in chain diagnostics
summarise_draws(stan2.3, default_convergence_measures()) |> 
  summarise(across(.cols = rhat:ess_tail, .fns = median))
summarise_draws(stan2.4, default_convergence_measures()) |> 
  summarise(across(.cols = rhat:ess_tail, .fns = median))
```

Speaking of comparison, it's now time to show our `stan()`-based `stan2.4` model contains identical results for the HMC draws as the `brm()` benchmark `brm2`. As a first step, we extract and save the posterior draws with `as_draws_df()`.

```{r}
draws_brm2    <- as_draws_df(brm2)
draws_stan2.4 <- as_draws_df(stan2.4)

# Check the structures of the objects
glimpse(draws_brm2)
glimpse(draws_stan2.4)
```

Before we compare the two, we'll need to rename two of the columns in `draws_stan2.4`.

```{r}
draws_stan2.4 <- draws_stan2.4 |> 
  rename(b_Intercept = `b[1]`,
         b_height = `b[2]`)
```

Now that the two draws objects have the same column names, we can compare them directly with `all.equal()`.

```{r}
all.equal(draws_brm2, draws_stan2.4)
```

Yes, their posterior draws are exactly the same. Success!

## Wrap up

In this post we fit a single-level Gaussian model with a single continuous predictor with **brms**, and then fit the equivalent model with **rstan**. To replicate the **brms** results exactly, we expanded our Stan syntax skills to use model matrices, parameter vectors, and the `normal_id_glm_lpdf()` function for the likelihood. In the [next post](https://solomonkurz.netlify.app/blog/2025-07-17-learn-stan-with-brms-part-iii/) we grapple with how **brms** mean-centers predictors, and what that means for setting priors and for the underlying Stan code.

Happy coding, friends!

## Thank the reviewers

I'd like to publicly acknowledge and thank

* [Riccardo Fusaroli](https://pure.au.dk/portal/en/persons/fusaroli%40cc.au.dk) and
* [Stephen J. Wild](https://sjwild.github.io/)

for their kind efforts reviewing the draft of this post. Go team!

Do note the final editorial decisions were my own, and I do not think it would be reasonable to assume my reviewers have given blanket endorsements of the current version of this post.

## Session information

```{r}
sessionInfo()
```

```{r, fig.width = 8, fig.height = 4, echo = F, eval = F}
xmax <- max(d$height) |> ceiling()

data.frame(height = 0:xmax) |> 
  add_epred_draws(brm2) |> 
  rename(weight = .epred) |> 
  
  ggplot(aes(x = height, y = weight)) +
  stat_lineribbon(.width = 0.95,
                  color = "steelblue4", fill = "steelblue1", linewidth = 1/2) +
  geom_point(data = d,
             size = 3/4) +
  theme(panel.grid = element_blank())

ggsave("height-featured.jpg", width = 3.6, height = 3.5, units = "in")
```

## References 

