---
title: Learn Stan with brms, Part I
subtitle: "y ~ 1"
author: A. Solomon Kurz
date: '2025-07-07'
excerpt: "In this series, we'll learn how to fit models with rstan, by parsing the Stan code underlying models fit with brms. In this first post, we start with a simple intercept-only Gaussian model."
tags:
  - Bayesian
  - brms
  - gaussian
  - GLM
  - rstan
  - stan
draft: false
layout: single
featured: no
bibliography: /Users/solomonkurz/Dropbox/blogdown/content/blog/my_blog.bib
biblio-style: apalike
csl: /Users/solomonkurz/Dropbox/blogdown/content/blog/apa.csl  
link-citations: yes
---

## Stan code can be hard

Though I'm an ardent **brms** user, I've been learning more about Stan for another project. One of the ways you can learn more about Stan is by reviewing the Stan code underlying your **brms** models. But I'm guessing that if you've ever tried that, you discovered Paul Bürkner's Stan code can be challenging to read. The purpose of this post, and the next couple posts to come, is to help demystify the Stan code from a few simple **brms** models.

### I make assumptions.

For this series, you'll want to be familiar with Bayesian single-level regression. We'll be sticking with Gaussian models for this series. For Bayesian regression beginners, I recommend the texts by Gelman and colleagues [-@gelmanRegressionOtherStories2020], Kruschke [-@kruschkeDoingBayesianData2015], or McElreath [-@mcelreathStatisticalRethinkingBayesian2015; -@mcelreathStatisticalRethinkingBayesian2020]. However, if applied Bayesian statistics is new to you, this probably isn't the right time to wade through this blog series.

All code will be in **R** [@R-base]. Data wrangling and plotting will rely heavily on the **tidyverse** [@R-tidyverse; @wickhamWelcomeTidyverse2019] and **ggdist** [@R-ggdist]. Bayesian models will be fit with **brms** [@R-brms; @burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018] and **rstan** [@R-rstan].^[Yes all you **CmdStan** lovers, I said **rstan**. I prefer CRAN-hosted packages, when they're available. You're at liberty to follow along using **CmdStan** on your own computer if you'd like, though.] We will post process our models with help from packages such as  **posterior** [@R-posterior] and **tidybayes** [@R-tidybayes].

For the **rstan** package, do check out their website at [https://mc-stan.org/rstan/](https://mc-stan.org/rstan/). But since **rstan** is mainly an **R** interface for Stan, readers of this blog series will want to bookmark the resources at the Stan website [https://mc-stan.org/docs/](https://mc-stan.org/docs/). Specifically, the [*Stan Functions Reference*](https://mc-stan.org/docs/functions-reference/) [@standevelopmentteamStanFunctionsReference2024], [*Stan Reference Manual*](https://mc-stan.org/docs/reference-manual/) [@standevelopmentteamStanReferenceManual2024], and the [*Stan User's Guide*](https://mc-stan.org/docs/stan-users-guide/) [@standevelopmentteamStanUserGuide2024]. For ongoing education and debugging insights, I also recommend setting up a profile on the Stan Forums at [https://discourse.mc-stan.org/](https://discourse.mc-stan.org/).

Load the primary **R** packages.

```{r, warning = F, message = F}
library(tidyverse)
library(brms)
library(rstan)
library(tidybayes)
library(ggdist)
library(posterior)
```

### We need data.

The data were made with a custom function, which you can learn about in [the appendix][appendix].

```{r}
load(file = "data/d.rda")
```

The data are $N = 100$ cases of height and weight values, simulated based on US adult benchmarks from a few decades ago. The units are inches and pounds.

```{r}
glimpse(d)
```

Here's what they look like in a scatter plot.

```{r, fig.width = 4.5, fig.height = 3}
d |> 
  ggplot(aes(x = height, y = weight)) +
  geom_point() +
  theme(panel.grid = element_blank())
```

For this post and the next two to come, we will take `weight` as our response variable^[I usually prefer to use the term *criterion variable*. As Bürkner tends to use the term *response variable* in his Stan code, it's probably best to just follow his lead to avoid any confusion.].

## Focal model

To start this series off, we will fit the simple intercept-only model

\begin{align}
\text{weight}_i & \sim \operatorname{Normal}(\mu, \sigma) \\
\mu & \sim \operatorname{Normal}(160, 25) \\
\sigma &  \sim \operatorname{Exponential}(0.04),
\end{align}

where `weight` values vary across $i$ people, and they are modeled as normally distributed with a simple population mean $\mu$ and standard deviation $\sigma$. In terms more common among regression models, you could also write $\mu = \beta_0$, the unconditional intercept. The last two rows are the priors, for which I'll provide some detail next.

The data were simulated based on US adult norms from a few decades ago, and the cases are an even mixture of men and women. We like our heavy dinners and soft drinks here in the US, and so our average weight tends to be higher than other populations [e.g., @owid2025obesity]. Based on self-reports from US adults from the early 1990s, their average weight at that time was around 160 pounds [@saad2021americans], which is why I've centered the prior for $\mu$ at 160. Setting the scale of the prior at 25 is a way to make it weak and uncertain on the scale of the data.

For the standard deviation of the `weight` distribution, I'm using the simple exponential prior, as recommended in @mcelreathStatisticalRethinkingBayesian2020. The mean of the exponential distribution is the reciprocal of its single parameter $\lambda$, and if we make a liberal guess the standard distribution is 25, that would mean $\lambda = 1/25 = 0.04$.

You, of course, don't have to use my priors. If you think they're terrible, great! Set your own and see what happens as you follow along.^[With hate in your heart and contempt in your eyes.] You'll probably learn more if you do.

## **brms** workflow

I'm assuming readers of this series are already familiar with the **brms** basics. If you're not, I've published several applied books on how to use **brms**, which you can find listed [here](https://solomonkurz.netlify.app/book/). However, the last two subsections in this section may be new to even to intermediate **brms** users.

### Fit the model with `brm()`.

We can fit this model with **brms** like so.^[Notice how I'm using the `file` argument to save my model as an external file. For this to work smoothly on your computer, you'll want to make a `file` folder in your working directory before hitting execute. You could also just remove that argument to avoid any file-saving complications. For more, execute `?brm` in your console to read the technical documentation.]

```{r brm1}
brm1 <- brm(
  data = d,
  family = gaussian,
  weight ~ 1,
  prior = prior(normal(160, 25), class = Intercept) +
    prior(exponential(0.04), class = sigma),
  cores = 4, seed = 1,
  file = "fits/brm1")
```

Check the parameter summary.

```{r}
print(brm1)
```

Our priors were a little high for $\mu$, and a little low for $\sigma$. But they were both well in the ballpark, so I'd call this a success. To get a better sense, here's a plot of the parameter posteriors (blue) superimposed on their prior distributions (gray).

```{r, fig.width = 8, fig.height = 3.25, warning = FALSE}
c(prior(normal(160, 25), class = Intercept),
  prior(exponential(0.04), class = sigma)) |> 
  parse_dist() |> 

  ggplot(aes(y = class)) +
  stat_slab(aes(dist = .dist_obj),
            fill = "gray80") +
  stat_histinterval(data = as_draws_df(brm1) |> 
                      pivot_longer(cols = sigma:Intercept, names_to = "class"),
                    aes(x = value),
                    .width = 0.95, color = "steelblue4", fill = "steelblue1") +
  scale_y_discrete(NULL, expand = expansion(mult = 0.1)) +
  xlab("parameter space") +
  theme(panel.grid = element_blank())
```

### Extract the Stan code with `stancode()`. {#stancode}

We can extract the Stan code underlying our `brm()` model with the `stancode()` function.

```{r}
stancode(brm1)
```

Even for a model without predictors, there's a lot going on in that code. Fortunately, there are simpler ways to specify a model like this in Stan, and we'll practice those first before building up to more sophisticated code like this. So if you're intimidated already, great! This is the blog post for you.

### Extract the Stan data with `standata()`.

Though we usually use data frames to input our data to **brms**, Stan expects data in a different format. To see the data **brms** actually passed onto Stan, we can use the `standata()` function.

```{r}
standata(brm1) |> 
  str()
```

Surprised? We'll explain the major features of that output below, and in the next two posts to come.

## **rstan** workflow

In this section I'll be pulling liberally from @kurz2025sr2rstan, [Chapter 1](https://solomon.quarto.pub/sr2rstan/01.html). Interested readers should go there for many more examples of Stan models fit to familiar data sets, especially if you like McElreath's books.

We will be fitting six **rstan** versions of the `brm1` model. We will start simple, and slowly build up our code until it matches all the major features from the `stancode()` output above. Though all of our **rstan**  models will be similar to the one from **brms**, the sixth and final model will match it exactly. The remaining subsections are organized around the six models, and the first is a massive **rstan** introductory with subsections of its own.

### `stan1.1` and the **rstan** newbie bootcamp.

Readers familiar with **rstan** might want to lightly browse through this section, and take whatever they may need. **brms** users new to **rstan** should probably work through it slowly, and spend some time with the resources cited within.

#### Stan likes data in lists.

Many model-fitting functions in **R** allow for a variety of data types, including data frames, lists, free-floating vectors, and so on. *Stan expects data in lists*, and the primary model-fitting functions in the **rstan** package generally expect data lists, too. The **tidybayes** package includes a `compose_data()` function, which makes it easy to convert data frames into the list format we need. Here's what this can look like for our use case.

```{r}
stan_data <- d |> 
  select(weight) |> 
  compose_data()

# What?
str(stan_data)
```

The `compose_data()` function automatically added a scalar value `n`, which defines the number of rows in the original data frame. As we will see, scalar values defining various dimensions are important for the kind of syntax we use with **rstan**. There were several in the data from `standata(brm1)` above, but we only need `n` to start.

#### `model_code` and its blocks.

Stan programs are organized into a series of program blocks, and those blocks are saved as a character string for the primary **rstan** functions.^[Technically, there are other ways to save the blocks, such as external files. Here we just rely on the character string method.] Following the schematic in the [*Program Blocks*](https://mc-stan.org/docs/reference-manual/blocks.html) section in the *Stan Reference Manual*, those blocks are:

```default
functions {
  // ... function declarations and definitions ...
}
data {
  // ... declarations ...
}
transformed data {
   // ... declarations ... statements ...
}
parameters {
   // ... declarations ...
}
transformed parameters {
   // ... declarations ... statements ...
}
model {
   // ... declarations ... statements ...
}
generated quantities {
   // ... declarations ... statements ...
}
```

These blocks must always be in this order, but not all Stan programs need all of the blocks. For example, our first version of the model only requires the `data`, `parameters`, and `model` blocks. Here's what they look like, saved as a character string named `model_code_1.1`.

```{r}
model_code_1.1 <- '
data {
  int<lower=1> n;    // Total number of observations
  vector[n] weight;  // Response variable
}
parameters {
  real mu;              // Unconditional mean (aka alpha, b_0, or beta_0)
  real<lower=0> sigma;  // Dispersion parameter
}
model {
  weight ~ normal(mu, sigma);  // Likelihood
  mu ~ normal(160, 25);        // Priors
  sigma ~ exponential(0.04);
}
'
```

In Stan program blocks, each line ends with a `;`. As you can see in the `model` block, we can annotate the code with `//` marks. We'll focus on each of the three program blocks in the subsections below.

#### Check the `data` block.

Data declarations are a *big deal* with Stan. We'll cover a few of the fine points in this section, but the primary information source is the [*Data Types and Declarations*](https://mc-stan.org/docs/reference-manual/types.html) section of the *Stan Reference Manual*. At some time or another, you will want to study that material.

Here's a focused look at our `data` block.

```default
data {
  int<lower=1> n;    // Total number of observations
  vector[n] weight;  // Response variable
}
```

All the data required by the model must be declared in the `data` block. An important exception is if you define a transformed version of any of your data in the `transformed data` block, and we will see examples of that strategy later in the third post of this series. One generally defines one variable per line, but it is possible to declare multiple variables in a single line (see [here](https://mc-stan.org/docs/reference-manual/types.html#declaring-multiple-variables-at-once)).

It's important to understand that Stan programs are read in order, top to bottom. Thus the first line in a `data` block often defines a scalar value we later use to define a data dimension. Recall how earlier we learned the `compose_data()` function automatically makes an `n` scalar, which defines the number of rows in the original data set. If you have more than one scalar value for your model, it's generally a good idea to declare them all at the top.

With the syntax of `int` in this first line, we have declared that `n` scalar an integer. Stan can accept two primitive number types, which are `int` for integers and `real` for continuous numbers. There are other special types, such as `complex`, but I believe we will just be using `int` and `real` in this blog series.

Some values have *constraints*, and we have declared `1` to be the lowest integer value allowed for our `n` scalar by the `<lower=1>` syntax in our first line. Technically, we didn't need to define this constraint for this example; the model will fit fine without it. In this case, the constraint will serve as a error check to make sure we have a least one case in our data. If you [look back][Extract the Stan code with `stancode()`.], you'll see Bürkner also used this convention in his Stan code.

By the second line `vector[n] weight`, we have defined our `weight` response values as a vector of length `n`. Stan supports several data types, such as scalars, vectors, matrices, and arrays (see [here](https://mc-stan.org/docs/reference-manual/types.html#overview-of-data-types)). Though sequences of real values (such as `weight`) can be declared in vectors or arrays, sequences of integers go into arrays. If you try to declare a sequence of integer values as a vector, Stan will return an error. In my experience, properly juggling vectors and arrays has been a major source of frustration. Stan is picky, friends. Respect the data types.

#### Our `parameters` block.

A good initial place to learn the technical details for the `parameters` block is in the [*Program block: `parameters`*](https://mc-stan.org/docs/reference-manual/blocks.html#program-block-parameters) section of the *Stan Reference Manual*. Otherwise you can glean a lot of applied insights from the [*Regression Models*](https://mc-stan.org/docs/stan-users-guide/regression.html) section of the *Stan User's Guide*.

Here's a focused look at our `parameters` block.

```default
parameters {
  real mu;              // Unconditional mean (aka alpha, b_0, or beta_0)
  real<lower=0> sigma;  // Dispersion parameter
}
```

Our unconditional mean parameter `mu` is declared as an unconstrained `real` value. Parameters can have constraints, and our syntax of `real<lower=0>` placed a lower boundary of zero for our $\sigma$ parameter. It's also possible to place upper boundaries, such as `<lower=0, upper=1>` for proportions. Though we don't have any here, you can declare vectors and matrices of parameters too.

#### Our `model` block. {#sec-Our-model-block}

In the above sections were we detailed our `data` and `parameters` blocks, all of their contents were what are called *declarations*.
Whereas `model` blocks do allow for declarations, they also allow for *statements*. In short, declarations let you name data elements and model parameters, and statements tell Stan how the parameters should be computed. A great place to learn all about statements is the [*Statements*](https://mc-stan.org/docs/reference-manual/statements.html) section of the *Stan Reference Manual*. You might also read the brief [*Program block: `model`*](https://mc-stan.org/docs/reference-manual/blocks.html#program-block-model) section of the *Stan Reference Manual*, or soak in all the applied examples in the [*Regression Models*](https://mc-stan.org/docs/stan-users-guide/regression.html) section of the *Stan User's Guide*. When you're really ready to get serious, you could browse though pretty much the whole of the *Stan Functions Reference*.

Here's a focused look at our `model` block.

```default
model {
  weight ~ normal(mu, sigma);  // Likelihood
  mu ~ normal(160, 25);        // Priors
  sigma ~ exponential(0.04);
}
```

My preference is to define the likelihood first, and then add the priors. I've seen many examples of the reverse, and sometimes they're even shuffled all around. You pick whatever convention that makes sense for you and your collaborators, but I do recommend you choose a consistent format.

With the syntax of `weight ~ normal(...)` in our first line, we have defined the `weight` response variable as normally distributed, and as described with the two parameters named `mu` and `sigma`. Even though $\mu$ and $\sigma$ are the typical names for these parameters, we could have chosen different ones for the model. But had we wanted to do so, that would have required different code in the `parameters` block. For example, say we wanted to call our first parameter `beta0` and our second parameter `scale`, we could have done this:

```default
parameters {
  real beta0;
  real<lower=0> scale;
}
model {
  weight ~ normal(beta0, scale);
  beta0 ~ normal(160, 25);
  scale ~ exponential(0.04);
}
```

But anyways, this kind of likelihood syntax is often called the *distribution* syntax in Stan. It's perhaps the most simple, and I generally like it the best. But there are many others, and Bürkner's Stan code above uses an alternative we'll discuss more [below][target].

Notice that our prior lines follow a similar kind of syntax as our likelihood line for `weight`. Each prior line started with the parameter of interest on the left side, followed by the tilde `~`, and then concluded with a distribution. Though not shown here, it is also possible to assign vectors of parameters to a common prior.

#### HMC sampling with `stan()`.

To my eye, there are two basic ways to draw posterior samples from **rstan**. The simplest way is with `stan()`, and a handy alternative is to use the combo of `stan_model()` and `sampling()`. In this blog series we'll just practice with the `stan()` function. The two main arguments are the `data` argument, into which we insert our `stan_data`, and the `model_code` argument, into which we insert our `model_code_1.1` string with its `data`, `parameters`, and `model` block information. There are several other arguments with default settings you might want to change. For example, `stan()` automatically samples from four HMC chains by the default setting `chains = 4`, which I generally find reasonable. Though it by default samples from the four chains in sequence, we will instead sample from them in parallel by setting `cores = 4`. To make the results more reproducible, I will also set `seed = 1`.

```{r stan1.1, echo = F}
# save(stan1.1, file = "fits/stan1.1.rda")
load(file = "fits/stan1.1.rda")
```

```{r, eval = F}
stan1.1 <- stan(
  data = stan_data,
  model_code = model_code_1.1,
  # These settings are optional. 
  # Notice how they mirror the settings we used with `brm()`
  cores = 4, seed = 1)
```

```{r, echo = F, eval = F}
print(brm1)
```

We can get a model summary with `print()`. I usually set `probs = c(0.025, 0.975)` to streamline the output a bit.

```{r}
# print(brm1)
print(stan1.1, probs = c(0.025, 0.975))
```

The parameter summaries are pretty similar to those from `print(brm1)` above, and by the [end of the post][prior_only] we'll get the two to match identically.

Notice that unlike the with the `print()` output for the **brms** model, the output for our **rstan** model also included a line for the `lp__`.  That stands for *log posterior*, and one way we can extract that value from our **brms** model is with the `log_posterior()` function. As the `lp__` has a distribution, we'll just summarize it by its mean and 95% interval with `mean_qi()`.

```{r}
log_posterior(brm1) |> 
  mean_qi(Value)
```

Again, the `lp__` for our two models does not match up yet, but it will by the end. As to its interpretation, the `lp__` is beyond the scope of this project. You can learn more about it from [Stephen Martin](https://srmart.in/)'s [nice explanation](https://discourse.mc-stan.org/t/basic-question-what-is-lp-in-posterior-samples-of-a-brms-regression/17567/2) on the Stan Forums.

### `stan1.2` by adding a `target`. {#target}

One of the many ways our simple `model_code` differed from Bürkner's is he used a very different kind of syntax for the likelihood and prior distributions. Part of that difference is he also used a `transformed parameters` block. Here's a focused look at those parts of his code:

```default
transformed parameters {
  real lprior = 0;  // prior contributions to the log posterior
  lprior += normal_lpdf(Intercept | 160, 25);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  // likelihood including constants
  if (!prior_only) {
    // initialize linear predictor term
    vector[N] mu = rep_vector(0.0, N);
    mu += Intercept;
    target += normal_lpdf(Y | mu, sigma);
  }
  // priors including constants
  target += lprior;
}
```

Before you fixate on it, Bürkner slipped an `if` statement into his `model` block. We'll come back to that [later][prior_only]. For now, focus on the other features of the code.

For now, notice how his syntax is full of lines with `+=` operators. These follow the common pattern `parameter_name` `+=` `distribution_name` `_lpdf`. The `lpdf` suffix stands for *log probability density function*. Whereas the `y ~ normal` style of syntax drops constants during computation, the `target += normal_lpdf` style of syntax retains them. As a consequence, the latter tends to be a bit slower, but it's also more general. Importantly, the  `target += normal_lpdf` style of syntax is required for some post-processing steps, such as Bayes factors. Though we won't be computing Bayes factors in this series, it's important for a package like **brms** to use a style of syntax that's general enough to handle use cases where they arise. You can begin to learn more about this kind of syntax in the [*Unbounded Continuous Distributions* section](https://mc-stan.org/docs/functions-reference/unbounded_continuous_distributions.html#normal-distribution) of the *Stan Functions Reference*, and the [*Distribution statements*](https://mc-stan.org/docs/reference-manual/statements.html#sampling-statements.section) section of the *Stan Reference Manual*.

We should also clarify `target` is a reserved term in Stan. By *target*, we mean the log posterior density function of the model parameters, or the log of the posterior density. That's why in Bürkner's code above, as well as in our code below, `target` is defined as the combination of the likelihood line (`normal_lpdf(weight | ...)`) and the priors (`lprior`), because in Bayes's theorem the posterior distribution is proportional to the likelihood times the prior.

Before we discuss the the `transformed parameters` block, we can take an intermediary step by defining our `model_code_1.2` with the `target += normal_lpdf` style of syntax like so:

```{r}
model_code_1.2 <- '
data {
  int<lower=1> n;
  vector[n] weight;
}
parameters {
  real mu;
  real<lower=0> sigma;
}
model {
  target += normal_lpdf(weight | mu, sigma);  // Likelihood
  target += normal_lpdf(mu | 160, 25);        // Priors
  target += exponential_lpdf(sigma | 0.04);
}
'
```

The only other thing to notice about this syntax is we have set `target` on the left side of the `+=` operator, and we use the notation of `name | parameters` within the parenthesis on the right. Here's how to fit the model with `stan()`.

```{r stan1.2, echo = F}
# save(stan1.2, file = "fits/stan1.2.rda")
load(file = "fits/stan1.2.rda")
```

```{r, eval = F}
stan1.2 <- stan(
  data = stan_data,
  model_code = model_code_1.2,
  cores = 4, seed = 1)
```

Check the model summary.

```{r}
# print(brm1)
# print(stan1.1, probs = c(0.025, 0.975))
print(stan1.2, probs = c(0.025, 0.975))
```

The summaries for the `mu` and `sigma` rows are largely the same as before. But notice the `lp__` has dropped to a different range, and it now matches more closely to what we saw from `log_posterior(brm1)` above.

### `stan1.3` to define the `lprior`.

We're now ready to say a few words about the `transformed parameters` block. The `transformed parameters` block is a place you can define additional parameters not necessarily part of the primary model, but from which you'd like to sample during the HMC process. Focus again on that part of Bürkner's code.

```default
transformed parameters {
  real lprior = 0;  // prior contributions to the log posterior
  lprior += normal_lpdf(Intercept | 160, 25);
  lprior += exponential_lpdf(sigma | 0.04);
}
```

In the first line he defined the `lprior`, which he described as the `prior contributions to the log posterior` in the comment. This is to provide support for automatic prior/likelihood sensitivity analyses using functions from the **priorsense** package [@kallioinen2023priorsense]. I have not used that package myself, but since Bürkner is one of the co-authors, we should not be surprised he has built support for it into **brms**. By the second and third lines, we can see the `lprior` is the sum of the log of the two priors. 

Now look again at Bürkner's `model` block.

```default
model {
  // likelihood including constants
  if (!prior_only) {
    // initialize linear predictor term
    vector[N] mu = rep_vector(0.0, N);
    mu += Intercept;
    target += normal_lpdf(Y | mu, sigma);
  }
  // priors including constants
  target += lprior;
}
```

As the priors have already been defined in the `transformed parameters` block, we can now just refer to them with the compact code `target += lprior` within `model`. Here's what that can look like for us.

```{r}
model_code_1.3 <- '
data {
  int<lower=1> n;
  vector[n] weight;
}
parameters {
  real mu;
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;  // Prior contributions to the log posterior
  lprior += normal_lpdf(mu | 160, 25);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  target += normal_lpdf(weight | mu, sigma);  // Likelihood
  target += lprior;                           // Priors
}
'
```

Fit the model.

```{r stan1.3, echo = F}
# save(stan1.3, file = "fits/stan1.3.rda")
load(file = "fits/stan1.3.rda")
```

```{r, eval = F}
stan1.3 <- stan(
  data = stan_data,
  model_code = model_code_1.3,
  cores = 4, seed = 1)
```

Check the model summary.

```{r}
# print(brm1)
# print(stan1.1, probs = c(0.025, 0.975))
# print(stan1.2, probs = c(0.025, 0.975))
print(stan1.3, probs = c(0.025, 0.975))
```

Now our `print()` summary contains a line for that mysterious `lprior`. Even though the `print(brm1)` output didn't include a line for that quantity, it has been lurking in the shadows all along. For example, here it is with the `posterior_summary()` method.

```{r}
posterior_summary(brm1)
```

It's also there when you extract the posterior draws with `as_draws_df()`.

```{r}
as_draws_df(brm1) |> 
  glimpse()
```

### `stan1.4` adds an `Intercept`.

Take another at Bürkner's `parameters` and `model` blocks.

```default
parameters {
  real Intercept;  // temporary intercept for centered predictors
  real<lower=0> sigma;  // dispersion parameter
}
model {
  // likelihood including constants
  if (!prior_only) {
    // initialize linear predictor term
    vector[N] mu = rep_vector(0.0, N);
    mu += Intercept;
    target += normal_lpdf(Y | mu, sigma);
  }
  // priors including constants
  target += lprior;
}
```

Whereas we've been defining our parameter for $\mu$ `mu` within the `parameters` block, Bürkner named that term `Intercept` in his. Since this is his general term for the model intercept (what you might also call $\alpha$ or $\beta_0$), this is no great surprise. We're not ready in this first post to discuss what Bürkner meant with his cryptic comment `temporary intercept for centered predictors`, but that will take a center stage in third post of this series. But also look how he smuggled a `mu += Intercept` line in the middle of that `if` statement, and then he used `mu` in his likelihood line: `target += normal_lpdf(Y | mu, sigma)`. In the context of this simple model, this seems like a redundant tautology. But this too will play a very important role in the third post of this series, and we will begin to discuss it more fully in the second post. So for now, just let the tension build and practice fitting a model with this extra step.

```{r}
model_code_1.4 <- '
data {
  int<lower=1> n;
  vector[n] weight;
}
parameters {
  real Intercept;  // Temporary intercept for centered predictors
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(Intercept | 160, 25);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  // Initialize linear predictor term
  vector[n] mu = rep_vector(0.0, n);
  mu += Intercept;
  target += normal_lpdf(weight | mu, sigma);
  target += lprior;
}
'
```

Fit the updated model.

```{r stan1.4, echo = F}
# save(stan1.4, file = "fits/stan1.4.rda")
load(file = "fits/stan1.4.rda")
```

```{r, eval = F}
stan1.4 <- stan(
  data = stan_data,
  model_code = model_code_1.4,
  cores = 4, seed = 1)
```

Check the model summary.

```{r}
# print(brm1)
# print(stan1.1, probs = c(0.025, 0.975))
# print(stan1.2, probs = c(0.025, 0.975))
# print(stan1.3, probs = c(0.025, 0.975))
print(stan1.4, probs = c(0.025, 0.975))
```

This time the `print()` output is very similar to what we saw before, but technically those changes to the `parameters` and `model` blocks did change the numbers on the small scale.

### `stan1.5` adds a `b_Intercept`.

Now notice that at the bottom of his code, Bürkner added a `generated quantities` block.

```default
generated quantities {
  // actual population-level intercept
  real b_Intercept = Intercept;
}
```

Unlike with other blocks, the quantities in the `generated quantities` block are computed after the HMC sampling is complete. This is generally a good place to define parameters that are deterministic functions of other parameters. Though in the last version of the model we already had a case of that with the `mu += Intercept`, that particular bit of code will look different in some of the models to come in future posts, and it was for a different purpose which we'll explain when it's more relevant. At the moment, this `real b_Intercept = Intercept` line also seems like a useless tautology. For now, just roll with it and practicing adding it to the code.

```{r}
model_code_1.5 <- '
data {
  int<lower=1> n;
  vector[n] weight;
}
parameters {
  real Intercept;
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(Intercept | 160, 25);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  vector[n] mu = rep_vector(0.0, n);
  mu += Intercept;
  target += normal_lpdf(weight | mu, sigma);
  target += lprior;
}
generated quantities {
  // Actual population-level intercept
  real b_Intercept = Intercept;
}
'
```

Fit the model.

```{r stan1.5, echo = F}
# save(stan1.5, file = "fits/stan1.5.rda")
load(file = "fits/stan1.5.rda")
```

```{r, eval = F}
stan1.5 <- stan(
  data = stan_data,
  model_code = model_code_1.5,
  cores = 4, seed = 1)
```

Check the model summary.

```{r}
# print(brm1)
# print(stan1.1, probs = c(0.025, 0.975))
# print(stan1.2, probs = c(0.025, 0.975))
# print(stan1.3, probs = c(0.025, 0.975))
# print(stan1.4, probs = c(0.025, 0.975))
print(stan1.5, probs = c(0.025, 0.975))
```

If you compare this `print()` output to that of the previous model, you'll see the results are identical, with the sole exception this one includes a new row for `b_Intercept`. Given how we defined `b_Intercept` as a deterministic function of `b_Intercept`, it should be unsurprising the summaries for the two parameters are also identical here.^[Again, this looks like madness for this simple model, but it will become extremely important for the third where we grapple with the **brms** default behavior for predictors.]

Though we aren't ready to explain why we'd want output for both `b_Intercept` and `Intercept`, notice how they are both present in the `as_draws_df()` output for the original `brm()` model `brm1`.

```{r}
as_draws_df(brm1) |> 
  glimpse()
```

### `stan1.6` for `prior_only` and `if`. {#prior_only}

We're finally ready to address that mysterious `if` statement in Bürkner's `model` block. We also need to review the `data` block. Here's a focused look at both.

```default
data {
  int<lower=1> N;  // total number of observations
  vector[N] Y;  // response variable
  int prior_only;  // should the likelihood be ignored?
}
model {
  // likelihood including constants
  if (!prior_only) {
    // initialize linear predictor term
    vector[N] mu = rep_vector(0.0, N);
    mu += Intercept;
    target += normal_lpdf(Y | mu, sigma);
  }
  // priors including constants
  target += lprior;
}
```

If you execute `?brm` in your console, you can see from the documentation file that the `brm()` function contains a `sample_prior` argument. The default setting is `"no"`, which means that `brm()` typically only takes draws from the posterior distribution. If you instead set that value to `"yes"`, `brm()` will sample from both the posterior *and* the priors, and if you set it to `"only"` you will only get back prior draws. The `if` statement in the `model` block is a way to handle that.

In the `data` block we see the last line includes a integer value named `prior_only`, and the comment on the right reads: `should the likelihood be ignored?` In the case where we use the `brm()` default setting `sample_prior = "no"`, that value is a `0`. Here's how we can add it to our data list.

```{r}
stan_data <- d |> 
  select(weight) |> 
  compose_data(prior_only = as.integer(FALSE))

str(stan_data)
```

I don't know the history behind why the `sample_prior` argument ends up passing its information to a `prior_only` value in the data, but that's what we've got. When you pass that `0` value to the `if` statement in the `model` block, the program will sample from the posterior. But if you were to pass a `1` instead, it would only sample from the prior. Here's how to build that into our version of the code.

```{r}
model_code_1.6 <- '
data {
  int<lower=1> n;
  vector[n] weight;
  int prior_only;  // Should the likelihood be ignored?
}
parameters {
  real Intercept;
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(Intercept | 160, 25);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  // Likelihood including constants
  if (!prior_only) {
    // Initialize linear predictor term
    vector[n] mu = rep_vector(0.0, n);
    mu += Intercept;
    target += normal_lpdf(weight | mu, sigma);
  }
  target += lprior;
}
generated quantities {
  real b_Intercept = Intercept;
}
'
```

Fit the model.

```{r stan1.6, echo = F}
# save(stan1.6, file = "fits/stan1.6.rda")
load(file = "fits/stan1.6.rda")
```

```{r, eval = F}
stan1.6 <- stan(
  data = stan_data,
  model_code = model_code_1.6,
  cores = 4, seed = 1)
```

Check the model summary.

```{r}
# print(brm1)
# print(stan1.1, probs = c(0.025, 0.975))
# print(stan1.2, probs = c(0.025, 0.975))
# print(stan1.3, probs = c(0.025, 0.975))
# print(stan1.4, probs = c(0.025, 0.975))
# print(stan1.5, probs = c(0.025, 0.975))
print(stan1.6, probs = c(0.025, 0.975))
```

It's difficult to compare this `print()` output with the original `print(brm1)` output from the beginning of the post, but our `stan1.6` model returned identical results. It'll be easier to compare the two directly by their `as_draws_df()` output. First we compute and save.

```{r}
draws_brm1    <- as_draws_df(brm1)
draws_stan1.6 <- as_draws_df(stan1.6)

# Check the structures of the objects
glimpse(draws_brm1)
glimpse(draws_stan1.6)
```

Before we compare the two, we'll need to adjust the column ordering for `draws_stan1.6`.

```{r}
draws_stan1.6 <- draws_stan1.6 |> 
  select(b_Intercept, sigma, Intercept, everything())
```

Now that the two draws objects follow the same structure, we can compare them directly with `all.equal()`.

```{r}
all.equal(draws_brm1, draws_stan1.6)
```

Yes, the posterior draws are exactly the same for `brm1` fit with **brms** and `stan1.6` fit with **rstan**. We did it!

We could double check by comparing their parameter summaries with `summarise_draws()`.

```{r}
summarise_draws(draws_brm1)
summarise_draws(draws_stan1.6)
```

Yep, their summaries are the same.

## Wrap up

For all you readers new to **rstan**: Congratulations on fitting your first `stan()` models! For all you more advanced readers who are feeling surly and restless, just chill; we've got nerdier stuff coming your way soon. In the next post, we expand our skill sets by adding a predictor variable to the model, and learn to grapple with matrix notation and more.

Happy coding, friends!

## Thank the reviewers

I'd like to publicly acknowledge and thank

* [Riccardo Fusaroli](https://pure.au.dk/portal/en/persons/fusaroli%40cc.au.dk) and
* [Stephen J. Wild](https://sjwild.github.io/)

for their kind efforts reviewing the draft of this post. Go team!

Do note the final editorial decisions were my own, and I do not think it would be reasonable to assume my reviewers have given blanket endorsements of the current version of this post.

## Session information

```{r}
sessionInfo()
```

## Appendix {#appendix}

Here is the code for the `height_weight_generator()` function. This is a light reformatting of the   `HtWtDataGenerator()` function from @kruschkeDoingBayesianData2015. The parameters in the function are based on @BrainardB1992.

```{r}
height_weight_generator <- function(n, male_prob = 0.5, seed = 47405) {

  # Male MVN distribution parameters
  height_male_mu <- 69.18;    height_male_sd <- 2.87
  log_weight_male_mu <- 5.14; log_weight_male_sd <- 0.17
  male_rho <- 0.42
  male_mean <- c(height_male_mu, log_weight_male_mu)
  male_sigma <- matrix(c(height_male_sd^2, male_rho * height_male_sd * log_weight_male_sd,
                         male_rho * height_male_sd * log_weight_male_sd, log_weight_male_sd^2),
                       nrow = 2)

  # Female cluster 1 MVN distribution parameters
  height_female_mu_1 <- 63.11;    height_female_sd_1 <- 2.76
  log_weight_female_mu_1 <- 5.06; log_weight_female_sd_1 <- 0.24
  female_rho_1 <- 0.41
  female_mean_1 <- c(height_female_mu_1, log_weight_female_mu_1)
  female_sigma_1 <- matrix(c(height_female_sd_1^2, female_rho_1 * height_female_sd_1 * log_weight_female_sd_1,
                             female_rho_1 * height_female_sd_1 * log_weight_female_sd_1, log_weight_female_sd_1^2),
                           nrow = 2)

  # Female cluster 2 MVN distribution parameters
  height_female_mu_2 <- 64.36;    height_female_sd_2 <- 2.49
  log_weight_female_mu_2 <- 4.86; log_weight_female_sd_2 <- 0.14
  female_rho_2 <- 0.44
  female_mean_2 <- c(height_female_mu_2, log_weight_female_mu_2)
  female_sigma_2 <- matrix(c(height_female_sd_2^2, female_rho_2 * height_female_sd_2 * log_weight_female_sd_2,
                             female_rho_2 * height_female_sd_2 * log_weight_female_sd_2, log_weight_female_sd_2^2),
                           nrow = 2)

  # Female cluster probabilities
  female_prob_1 <- 0.46
  female_prob_2 <- 1 - female_prob_1

  # Randomly generate data values from those MVN distributions
  if (!is.null(seed)) set.seed(seed)
  datamatrix <- matrix(0, nrow = n, ncol = 3)
  colnames(datamatrix) <- c("male", "height", "weight")
  male_val <- 1; female_val <- 0  # Arbitrary coding values
  for (i in 1:n) {
    # Generate `sex`
    sex <- sample(c(male_val, female_val), size = 1, replace = TRUE, prob = c(male_prob, 1 - male_prob))
    if (sex == male_val) datum <- MASS::mvrnorm(n = 1, mu = male_mean, Sigma = male_sigma)
    if (sex == female_val) {
      female_cluster <- sample(c(1, 2), size = 1, replace = TRUE, prob = c(female_prob_1, female_prob_2))
      if (female_cluster == 1) datum <- MASS::mvrnorm(n = 1, mu = female_mean_1, Sigma = female_sigma_1)
      if (female_cluster == 2) datum <- MASS::mvrnorm(n = 1, mu = female_mean_2, Sigma = female_sigma_2)
    }
    datamatrix[i, ] <- c(sex, round(c(datum[1], exp(datum[2])), digits = 1))
  }

  data.frame(datamatrix)
}
```

I made and saved the data for these first three posts like so.

```{r}
d <- height_weight_generator(n = 100, seed = 1)

save(d, file = "data/d.rda")
```

```{r, echo = F, eval = F}
ymin <- min(d$weight)  # 99.4
ymax <- max(d$weight)  # 227.5

as_draws_df(brm1) |> 
  slice_sample(n = 30) |> 
  expand_grid(weight = seq(from = ymin - 50, to = ymax + 50, length.out = 101)) |> 
  mutate(density = dnorm(x = weight, mean = b_Intercept, sd = sigma)) |> 
  
  ggplot(aes(x = weight)) +
  geom_histogram(data = d,
                 aes(y = after_stat(density)),
                 binwidth = 10, fill = "steelblue1") +
  geom_line(aes(y = density, group = .draw),
            alpha = 2/3, color = "steelblue4", linewidth = 1/3) +
  scale_y_continuous(NULL, breaks = NULL) +
  coord_cartesian(xlim = c(ymin - 40, to = ymax + 40)) +
  theme(panel.grid = element_blank())
  
ggsave("height-featured.jpg", width = 3.6, height = 3.5, units = "in")
```

## References

