---
title: Learn Stan with brms, Part III
subtitle: "y ~ 1 + x"
author: A. Solomon Kurz
date: today
excerpt: "In this third post we explore different ways to fit a model with a mean-centered predictor, what this means for how we set our priors with **brms**, and what this all means for the underlying Stan code. Along the way we practice with the `transformed data` block, and get fancy with the model matrix."
tags:
  - Bayesian
  - brms
  - gaussian
  - GLM
  - regression
  - rstan
  - stan
draft: false
layout: single
featured: no
bibliography: /Users/solomonkurz/Dropbox/blogdown/content/blog/my_blog.bib
biblio-style: apalike
csl: /Users/solomonkurz/Dropbox/blogdown/content/blog/apa.csl  
link-citations: yes
nocite: |  # This adds references not cited in the prose
  @R-base, @R-tidyverse, @wickhamWelcomeTidyverse2019, @R-ggdist, @R-brms, @burknerBrmsPackageBayesian2017, @burknerAdvancedBayesianMultilevel2018, @R-rstan, @R-posterior, @R-tidybayes
---

## Center the $x$

In the last post, we fit a model with a single continuous predictor using the very non-default **brms** syntax `bf(y ~ 1 + x, center = FALSE)`. We also had some difficulty settling on a sensible prior for the intercept in this context, given how all our predictor values were far above zero. In this post we'll explore different ways to fit the model with a mean-centered predictor, what this means for how we set our priors with **brms**, and what this all means for the underlying Stan code. Along the way we practice with the `transformed data` block, and get fancier with the model matrix.

Load the primary **R** packages and the data file.

```{r, eval = F}
# Packages
library(tidyverse)
library(brms)
library(rstan)
library(tidybayes)
library(ggdist)
library(posterior)
library(patchwork)

# Data
load(file = "data/d.rda")

# Review
glimpse(d)
```

```{r, warning = F, message = F, echo = F}
# Packages
library(tidyverse)
library(brms)
library(rstan)
library(tidybayes)
library(ggdist)
library(posterior)
library(patchwork)

# Data
load(file = "/Users/solomonkurz/Dropbox/blogdown/content/blog/2025-07-07-learn-stan-with-brms-part-i/data/d.rda")

# Review
glimpse(d)
```

We continue with the height and weight data. For the code underlying the data, review the [appendix](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#appendix) from the first post.

## Focal model

In this post we will fit the univariable model

$$
\begin{align}
\text{weight}_i & \sim \operatorname{Normal}(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 \text{height}_i \\
\beta_0 & \sim \mathord ? \\
\beta_1 & \sim \operatorname{Normal}(5, 2.5) \\
\sigma &  \sim \operatorname{Exponential}(0.04),
\end{align}
$$

where `weight` values vary across $i$ people, and they are modeled as normally distributed with a conditional mean $\mu_i$ and [residual] standard deviation $\sigma$. The conditional mean is defined by an intercept $\beta_0$ and slope $\beta_1$ for the predictor variable `height`.

*That's the same model as last time*, you say. Well, yes and no. In another sense, we'll also be fitting the model

$$
\begin{align}
\text{weight}_i & \sim \operatorname{Normal}(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 \left ( \text{height}_i - \overline{\text{height}} \right ) \\
\beta_0 & \sim \mathord ? \\
\beta_1 & \sim \operatorname{Normal}(5, 2.5) \\
\sigma &  \sim \operatorname{Exponential}(0.04),
\end{align}
$$

where $\overline{\text{height}}$ is the sample mean for `height`, and $(\text{height}_i - \overline{\text{height}})$ is mean-centered `height`. *But wait; which one are we fitting?* you ask. In a way, we'll be fitting both.

As to the priors, you may have noticed two things. First, there's a big question mark by the intercept $\beta_0$, which is my way of indicating we're not ready to discuss that prior, yet. But otherwise you may have also noticed the priors for $\beta_1$ and $\sigma$ are the same as in the last post. Before we can fully discuss the prior for $\beta_0$, we need to discuss how **brms** handles mean-centering predictors.

## **brms** workflow

### `brm()` automatically mean centers your predictors

Read that header again. Unless you tell it otherwise, *the* `brm()` *function automatically mean centers your predictors*. After mean centering the predictors, it then fits the model with those centered predictors as the inputs. This has no effect on the coefficients for the predictors themselves (the regression slopes), but is has a profound effect on the "intercept." At this point, we should probably quote from the [*brms Reference Manual*](https://cran.r-project.org/web/packages/brms/brms.pdf) [@brms2024RM] at length. Here's the *Parameterization of the population-level intercept* section, in full:

> By default, the population-level intercept (if incorporated) is estimated separately and not as part of population-level parameter vector `b`. As a result, priors on the intercept also have to be specified separately. Furthermore, to increase sampling efficiency, the population-level design matrix `X` is centered around its column means `X_means` if the intercept is incorporated. This leads to a temporary bias in the intercept equal to `<X_means, b>`, where `<,>` is the scalar product. The bias is corrected after fitting the model, but be aware that you are effectively defining a prior on the intercept of the centered design matrix not on the real intercept. You can turn off this special handling of the intercept by setting argument `center` to `FALSE`. For more details on setting priors on
population-level intercepts, see `set_prior`.
>
> This behavior can be avoided by using the reserved (and internally generated) variable `Intercept`. Instead of `y ~ x`, you may write `y ~ 0 + Intercept + x`. This way, priors can be defined on the real intercept, directly. In addition, the intercept is just treated as an ordinary population-level effect and thus priors defined on `b` will also apply to it. Note that this parameterization may be less efficient than the default parameterization discussed above.

As indicated in the first paragraph, this has consequences for how you set your priors. From the `set_prior` section, we read:

> In case of the default intercept parameterization (discussed in the 'Details' section of `brmsformula`), general priors on class `"b"` will *not* affect the intercept. Instead, the intercept has its own parameter class named `"Intercept"` and priors can thus be specified via `set_prior("<prior>", class = "Intercept")`. Setting a prior on the intercept will not break vectorization of the other population-level effects. Note that technically, this prior is set on an intercept that results when internally centering all population-level predictors around zero to improve sampling efficiency. On this centered intercept, specifying a prior is actually much easier and intuitive than on the original intercept, since the former represents the expected response value when all predictors are at their means. To treat the intercept as an ordinary population-level effect and avoid the centering parameterization, use `0 + Intercept` on the right-hand side of the model formula.

To understand what this all means in practice, consider a scenario where you'd like to fit the model `y ~ 1 + x`, where the predictor `x` is not mean centered (such as in our case with `height`). If you would like to set the prior for $\beta_0$ given a non-centered predictor, you can use one of these two strategies:

```default
# The `bf(center = FALSE)` method
brm(
  ...,
  bf(y ~ 1 + x, center = FALSE),
  prior = prior(normal(<mean>, <sd>), class = b, coef = Intercept),
  ...)

# The `0 + Intercept` method
brm(
  ...,
  y ~ 0 + Intercept + x,
  prior = prior(normal(<mean>, <sd>), class = b, coef = Intercept),
  ...)
```

The first strategy sets `center = FALSE` within the `bf()` function, and the second strategy uses the `0 + Intercept` syntax. In both cases, when you set the prior for $\beta_0$, it is of `class = b` and `coef = Intercept`. The first of these two options is what we did for `brm2` from the [last post](https://solomonkurz.netlify.app/blog/2025-07-13-learn-stan-with-brms-part-ii/#fit-the-model-with-brm). The relevant bit of code was:

```default
brm(
  ...,
  bf(weight ~ 1 + height, center = FALSE),
  prior = prior(normal(0, 100), class = b, coef = Intercept),
  ...)
```

However, if you would like to set the prior for $\beta_0$ as if you were using a mean-centered predictor and therefore $\beta_0$ was "the expected response value when all predictors are at their means," then you can use the simpler default syntax of:

```default
# The default method
brm(
  ...,
  y ~ 1 + x,
  prior = prior(normal(<mean>, <sd>), class = Intercept),
  ...)
```

But do keep in mind that the $\beta_0$ intercept under this prior *is not* the $\beta_0$ coefficient displayed in the `print()` summary. Rather, that $\beta_0$ coefficient is computed "after fitting the model," which we will demonstrate below with the `stan()` versions of the model. This will be our approach here. One of Bürkner's claims in the quotes above was "specifying a prior is actually much easier and intuitive [on this centered intercept] than on the original intercept, since the former represents the expected response value when all predictors are at their means." To get a sense of what that means for our use case, here's the model we'll fit:

$$
\begin{align}
\text{weight}_i & \sim \operatorname{Normal}(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 \left ( \text{height}_i - \overline{\text{height}} \right ) \\
\beta_0 & \sim \operatorname{Normal}(160, 25) \\
\beta_1 & \sim \operatorname{Normal}(5, 2.5) \\
\sigma &  \sim \operatorname{Exponential}(0.04),
\end{align}
$$

where even though we're entering `height` into the formula in `brm()`, we know that under the hood it's actually using $(\text{height}_i - \overline{\text{height}})$. The $\operatorname{Normal}(160, 25)$ prior for $\beta_0$ given the mean-centered predictor is the same prior we used for the unconditional mean in `brm1`, the model from the [first post](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#fit-the-model-with-brm). As a general strategy, whatever prior you would use for an unconditional mean, I suspect that would also be a good prior for a model conditional on mean-centered predictors.

### Fit the model with `brm()`

Here's how to fit the model with `brm()`. Note how since we are using the simple `y ~ 1 + x` syntax, the prior for $\beta_0$ is simply of `class = Intercept`.

```{r brm3}
brm3 <- brm(
  data = d,
  weight ~ 1 + height,
  prior = prior(normal(160, 25), class = Intercept) +
    prior(normal(5, 2.5), class = b) +
    prior(exponential(0.04), class = sigma),
  cores = 4, seed = 1,
  file = "fits/brm3")
```

Review the `print()` summary.

```{r}
print(brm3)
```

The summaries for $\beta_1$ and $\sigma$ are about the same as they were for `brm2` from the [second post](https://solomonkurz.netlify.app/blog/2025-07-13-learn-stan-with-brms-part-ii/#fit-the-model-with-brm). The `Intercept` line is a summary for the intercept had you fit the model with the non-mean centered version of the predictor `height`. There is no summary of the actual intercept `brm()` used internally to fit the model from `print()`, but we do still have access to that parameter. Behold the summary from the `posterior_summary()` function:

```{r}
posterior_summary(brm3)
```

Confusingly, the top row `b_Intercept` here is the same as the row called `Intercept` in the `print()` output. The new line called `Intercept` is the intercept for the model using mean-centered `height`. Notice how that value is about the same as the unconditional mean from `brm1` from the [first post](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#fit-the-model-with-brm),^[Though with a smaller posterior standard deviation, as we have now accounted for some of the variance with the predictor.] and that value is pretty close to where we centered our prior.

You also get the same parameter-name convention in functions like `as_draws_df()`.

```{r}
draws_brm3 <- as_draws_df(brm3)  # Compute and save
glimpse(draws_brm3)              # Display
```

The first column `b_Intercept` is the intercept had you fit the model with non-mean-centered `height`, and the fourth column `Intercept` is the parameter we actually set the prior on, the intercept given mean-centered `height`.

To further illustrate this whole confusing mess, let's make a plot.

```{r, fig.width = 8, fig.height = 4.5, warning = FALSE}
xmax <- max(d$height) |> ceiling()
ymax <- max(d$weight)
ymin <- as_draws_df(brm3) |> summarise(quantile(b_Intercept, 0.025)) |> pull()
m_x <- mean(d$height)
m_y <- mean(d$weight)

# Left
p1 <- data.frame(height = 0:xmax) |> 
  add_epred_draws(brm3) |> 
  rename(weight = .epred) |> 
  
  ggplot(aes(x = height, y = weight)) +
  geom_hline(yintercept = m_y, color = "white") +
  geom_vline(xintercept = m_x, color = "white") +
  stat_lineribbon(.width = 0.95,
                  color = "steelblue4", fill = "steelblue1", linewidth = 1/2) +
  geom_point(data = d) +
  coord_cartesian(ylim = c(ymin, ymax)) +
  theme(panel.grid = element_blank())

# Right
p2 <- draws_brm3 |> 
  pivot_longer(cols = contains("Intercept")) |> 
  
  ggplot(aes(y = value, x = name)) +
  geom_hline(yintercept = m_y, color = "white") +
  stat_pointinterval(.width = 0.95, 
               color = "steelblue4", size = 1/10) +
  scale_x_discrete(NULL, position = "top") +
  scale_y_continuous(NULL, breaks = NULL, limits = c(ymin, ymax)) +
  theme(axis.text.x = element_text(angle = 270, hjust = 1),
        panel.grid = element_blank())

# Combine, format, and display
p1 + p2 +
  plot_layout(widths = c(15, 1))
```

In the main panel on the left, you have the fitted line which looks a lot like the one from the [second post](https://solomonkurz.netlify.app/blog/2025-07-13-learn-stan-with-brms-part-ii/#fit-the-model-with-brm). The white lines intersecting in the middle of the data cloud mark off the sample means for `height` and `weight`. Now notice the skinny coefficient plot in the right panel. See the wide point interval toward the bottom? That marks of the version of $\beta_0$ called `b_Intercept` in the `posterior_summary()` and `as_draws_df()` output; that's the intercept had we fit the model using `height`. But notice the point interval for the `Intercept` coefficient. That's the one for the intercept with the model fit to mean-centered `height`, and that's the coefficient for which we set the prior `prior(normal(160, 25), class = Intercept)`.

### Real talk

In my opinion, this was easily the worst programming decision Bürkner made for **brms**. I hate it that when I enter `y ~ 1 + x` in the formula line, I'm asked to set the intercept prior based on `y ~ 1 + (x - mean(x))` as a default. It's inherently confusing, and the way that `print()` and `posterior_summary()` use different naming conventions for the intercept makes it worse. I'd bet a lot of **brms** users don't even understand this, which makes me suspect that several folks may have published scientific articles based on models fit using the wrong intercept priors. It's a disaster.

If you want to avoid all this, either set `center = FALSE`, or use the `0 + Intercept` syntax. This is what I do most of the time for my real-world projects. When I want to fit a model with mean-centered predictors (which I sometimes do), I prefer centering them myself. *But the* brms Reference Manual *said this was all to "increase sampling efficiency,"* you say. I do not care. Personally, I would much rather take a little longer during the sampling phase than risk incorrectly setting a prior out of syntax confusion.

However, since this is what `brms()` does, we can at least work to understand it further by diving into the underlying Stan code, and reproducing the results with `stan()`. But before we do, let's practice a little algebra.

### Compute `b_Intercept` from `Intercept`, by hand

We can compute $\beta_0$ using the equation

$$
\beta_0 = \beta_{0_\text{temporary}} - \overline{\text{height}} \beta_1
$$

where $\overline{\text{height}}$ is the sample mean for `height`, and $\beta_1$ is the regression coefficient for `height`. With this notation, I'm calling the intercept for `height` $\beta_0$, and the intercept for mean-centered `height` $\beta_{0_\text{temporary}}$. In the `as_draws_df()` output, these are `b_Intercept` and `Intercept`, respectively.

Here's what this algebra can look like in code. We'll plot the results.

```{r, fig.width = 6, fig.height = 3, warning = FALSE}
draws_brm3 |> 
  # Use the equation
  mutate(`Intercept - m_x * b_height` = Intercept - m_x * b_height) |> 
  pivot_longer(cols = c(b_Intercept, `Intercept - m_x * b_height`)) |> 
  
  # Plot
  ggplot(aes(x = value, y = name)) +
  stat_halfeye(.width = 0.95) +
  scale_y_discrete(NULL, expand = expansion(mult = 0.1)) +
  xlab(expression(beta[0])) +
  theme(panel.grid = element_blank())
```

Are they identical?

```{r}
draws_brm3 |> 
  mutate(b_Intercept_by_hand = Intercept - (m_x * b_height)) |> 
  summarise(all_equal = all.equal(b_Intercept, b_Intercept_by_hand))
```

Yes. We'll see more of that algebra in action in the Stan code below.

### Extract the Stan code with `stancode()`

Extract the Stan code underlying our `brm3` model with the `stancode()` function.

```{r}
stancode(brm3)
```

A lot of this stuff should look familiar at this point. But there are a few new features scattered throughout several of the blocks.

### Extract the Stan data with `standata()`

Investigate the data **brms** actually passed onto Stan with the `standata()` function.

```{r}
standata(brm3) |> 
  str()
```

We have seen all of these features before.^[Even the `Kc` scalar was actually present in the data from `brm1` from the [first post](https://solomonkurz.netlify.app/blog/2025-07-07-learn-stan-with-brms-part-i/#extract-the-stan-data-with-standata). In a [later section][`stan3.2` for new tricks and the final model] we'll clarify what it actually is.]

## **rstan** workflow

This time we will only fit two models. We'll start very simple with `stan3.1`, and finish by implementing all the new features with `stan3.2`.

### `stan3.1`, start as simple as one can

First, we define the data list with `compose_data()`.

```{r}
stan_data <- d |> 
  select(weight, height) |> 
  compose_data()

str(stan_data)
```

For this version of the data, we only need the `weight` and `height` values as vectors, and the only scalar necessary is `n`.

We can define the simplest version of the `model_code` like so.

```{r}
model_code_3.1 <- '
data {
  int<lower=1> n;
  vector[n] weight;
  vector[n] height;
}
transformed data {
  real mean_height = mean(height);            // Sample mean for the predictor
  vector[n] height_c = height - mean_height;  // Mean-centered version of the predictor
}
parameters {
  real beta0_c;  // Temporary intercept for centered predictors
  real beta1;
  real<lower=0> sigma;
}
model {
  weight ~ normal(beta0_c + beta1 * height_c, sigma);  // Likelihood
  beta0_c ~ normal(160, 25);                           // Priors
  beta1 ~ normal(5, 2.5);
  sigma ~ exponential(0.04);
}
generated quantities {
  // Compute the actual intercept for the model using the non-centered predictor
  real beta0 = beta0_c - mean_height * beta1;
}
'
```

Of note, we have two lines in our `transformed data` block. The first line computes the mean of the `height` vector, and saves that quantity as a real value named `mean_height`. The second line subtracts the `mean_height` value from each row in the `height` vector, and saves it as a new vector named `height_c`, the `_c` suffix for which stands for *centered*.

The `beta1` and `sigma` parameters in our `parameters` block are business-as-usual at this point. The first line, however, defines `beta0_c` as the "`Temporary intercept for centered predictors`."

Next focus on the `model` block. In the likelihood line we use the `height_c` vector as the predictor which is multiplied with `beta1`, and the `beta0_c` parameter serves as the intercept. The `normal(160, 25)` prior is applied to the temporary `beta0_c` intercept, and the other two priors are used in the usual way.

Finally notice the `generated quantities` block. There we use the equation from [above][Compute `b_Intercept` from `Intercept`, by hand] to compute `beta0`, which is the intercept for the model had it been fit with the *non*-centered vector `height`. Recall that because it is defined in the `generated quantities`, the `beta0` parameter is computed after the HMC samples are completed.

Fit the model with `stan()`.

```{r stan3.1, echo = F}
# save(stan3.1, file = "fits/stan3.1.rda")
load(file = "fits/stan3.1.rda")
```

```{r, eval = F}
stan3.1 <- stan(
  data = stan_data,
  model_code = model_code_3.1,
  cores = 4, seed = 1)
```

Now we check the model summary. Here we'll compare it directly with the `posterior_summary()` summary for `brm3`, as also shown above.

```{r}
# print(brm3)
posterior_summary(brm3) |> 
  round(digits = 2)
print(stan3.1, probs = c(0.025, 0.975))
```

The summaries overlap nicely. What the `stan3.1` output calls `beta0_c` and `beta0`, the `brm3` output calls `Intercept` and `b_Intercept`.

### `stan3.2` for new tricks and the final model

For the data list, add the `prior_only` value and the model matrix.

```{r}
mm3.2 <- model.matrix(data = d, object = ~height)

stan_data <- d |> 
  select(weight) |> 
  compose_data(k = ncol(mm3.2),
               kc = ncol(mm3.2) - 1,
               X = mm3.2,
               prior_only = as.integer(FALSE))

str(stan_data, give.attr = FALSE)
```

Note how we now define the total number of columns in the model matrix as `k`, and `kc` is one minus that value. We subtract one because the `kc` value is excluding the vector of `1`'s for the intercept, which we'll explain in more detail later.

Here's how to make the full version of the `model_code`.

```{r}
model_code_3.2 <- '
data {
  int<lower=1> n;
  int<lower=1> k;   // Number of beta coefficients in the linear model
  int<lower=1> kc;  // number of beta coefficients minus the intercept
  vector[n] weight;
  matrix[n, k] X;
  int prior_only;
}
transformed data {
  matrix[n, kc] Xc;    // Centered version of X without an intercept
  vector[kc] means_X;  // Column means of X before centering
  for (i in 2:k) {
    means_X[i - 1] = mean(X[, i]);
    Xc[, i - 1] = X[, i] - means_X[i - 1];
  }
}
parameters {
  vector[kc] b;    // Beta coefficients (minus the intercept)
  real Intercept;  // Temporary intercept for centered predictors
  real<lower=0> sigma;
}
transformed parameters {
  real lprior = 0;
  lprior += normal_lpdf(b | 5, 2.5);
  lprior += normal_lpdf(Intercept | 160, 25);
  lprior += exponential_lpdf(sigma | 0.04);
}
model {
  // Likelihood
  if (!prior_only) {
    target += normal_id_glm_lpdf(weight | Xc, Intercept, b, sigma);
  }
  target += lprior;
}
generated quantities {
  // Compute the actual intercept for the model using the non-centered predictor
  real b_Intercept = Intercept - dot_product(means_X, b);
}
'
```

We'll comment on the various blocks in order.

For the `data` block, the only new feature is the third line for the `kc` scalar, which we mentioned before is one less than `k`. The fifth line for the `X` matrix is the same as in the last post, and this is the third time we've seen the `prior_only` value.

We should probably take a focused look at the `transformed data` block.

```default
transformed data {
  matrix[n, kc] Xc;    // Centered version of X without an intercept
  vector[kc] means_X;  // Column means of X before centering
  for (i in 2:k) {
    means_X[i - 1] = mean(X[, i]);
    Xc[, i - 1] = X[, i] - means_X[i - 1];
  }
}
```

The first line defines a new matrix called `Xc`, which is the same as the `X` matrix, but without the first column for the intercept. Though the dimensions for the `Xc` matrix are declared in the first line, the actual contents are computed in the final line, to which we'll return in a moment. In the second line we declare `means_X` as a vector of length `kc`, which is one in our case. Though this notation is perhaps overkill for a single-predictor model, it appears designed to scale up for cases with multiple predictors. Then we arrive at the `for` loop, which is perhaps the heart of the block. The first line in the `for` loop computes the mean for the second through `k` columns in the `X` matrix, and saves those values in the `means_X` vector. Then the second line in the loop defines the contents of the new `Xc` matrix as the mean-centered versions of the non-intercept columns of `X`.

Now focus on the `parameters` block.

```default
parameters {
  vector[kc] b;    // Beta coefficients (minus the intercept)
  real Intercept;  // Temporary intercept for centered predictors
  real<lower=0> sigma;
}
```

The first line declares all non-intercept parameters as a vector of dimension `kc`, named `b`. In our case this is just the slope $\beta_1$. The second line defines the `Temporary intercept for centered predictors`. That is, this is the intercept for the model fit using the mean-centered vector for `height`. Earlier we called this `beta0_c`, but this time we just adopt Bürkner's name `Intercept`. The `sigma` line remains as it has in all models before.

We have no new features in the `transformed parameters` block. It's used to compute the `lprior`, much as in all previous instances.

The `model` block deserves a focused look.

```default
model {
  // Likelihood
  if (!prior_only) {
    target += normal_id_glm_lpdf(weight | Xc, Intercept, b, sigma);
  }
  target += lprior;
}
```

The `if` statement is for the `prior_only` issue which allows `brm()` to sample from the posterior and/or the prior. In our case we'll be sampling from the posterior. The bottom line for `target += lprior` is the same as in previous cases, too. However, in the heart of the `if` statement we see new features within the `normal_id_glm_lpdf()` function. Notice we have input the `Xc` matrix into the $x$ argument, rather than the full `X` matrix. That is, our data matrix *does not* include the constant vector of `1`'s. To accommodate this, notice how we input `Intercept` into the $\alpha$ argument. Recall that in model `stan2.4` from the [last post](https://solomonkurz.netlify.app/blog/2025-07-13-learn-stan-with-brms-part-ii/#stan24-with-the-normal_id_glm_lpdf-function), we input a `0` to the $\alpha$ argument at the same time we input the entire `X` matrix into the $x$ argument. As for the $\beta$ argument, the input is nominally the same as before: `b`. However, whereas `b` was a vector of parameters including the intercept in the last post, the `b` vector only includes non-intercept $\beta$ coefficient for this case. The $\sigma$ argument remains the same as before.

Finally, we take a focused look at the `generated quantities` block.

```default
generated quantities {
  // Compute the actual intercept for the model using the non-centered predictor
  real b_Intercept = Intercept - dot_product(means_X, b);
}
```

Stan uses specialized operators for certain input types. In the simple model `stan3.1` above, we just multiplied the `mean_height` scalar with the single real parameter `beta1` using the `*` operator. In this case, `means_X` and `b` are both vectors, and therefore they require the `dot_product()` operator, which can multiply two vectors. For documentation on this class of operators, see the [*Dot products and specialized products*](https://mc-stan.org/docs/functions-reference/matrix_operations.html#dot-products-and-specialized-products) section of the *Stan Functions Reference* [@standevelopmentteamStanFunctionsReference2024]. Remember friends, Stan is *very* picky about data types.

Okay, let's fit the model.

```{r stan3.2, echo = F}
# save(stan3.2, file = "fits/stan3.2.rda")
load(file = "fits/stan3.2.rda")
```

```{r, eval = F}
stan3.2 <- stan(
  data = stan_data,
  model_code = model_code_3.2,
  cores = 4, seed = 1)
```

Now we check the model summary.

```{r}
# summary(brm3)
# print(stan3.1, probs = c(0.025, 0.975))
print(stan3.2, probs = c(0.025, 0.975))
```

To compare these results with the `brm()` benchmark, we first extract and save the posterior draws with `as_draws_df()`.

```{r}
draws_stan3.2 <- as_draws_df(stan3.2)

# Check the structures of the two draws objects
glimpse(draws_brm3)
glimpse(draws_stan3.2)
```

Before we compare the two, we need to rename the `b[1]` column and adjust the order of the columns in `draws_stan3.2`.

```{r}
draws_stan3.2 <- draws_stan3.2 |> 
  rename(b_height = `b[1]`) |> 
  select(b_Intercept, b_height, sigma, everything()) 
```

Now that the two draws objects have the same column names and column ordering, we can compare them directly with `all.equal()`.

```{r}
all.equal(draws_brm3, draws_stan3.2)
```

Yes, their posterior draws are exactly the same. Success!

## Wrap up

Similar to the [last](https://solomonkurz.netlify.app/blog/2025-07-13-learn-stan-with-brms-part-ii/), in this post we fit a single-level Gaussian model with a single continuous predictor with **brms**, and then fit the equivalent model with **rstan**. But this time we grappled with how **brms** automatically mean-centers predictors, for what this means for how you set the prior for the intercept, and what the intercept even means from a conceptual perspective. We further developed our Stan skills with model matrices and the `normal_id_glm_lpdf()` function, and we introduced data transformations in the `transformed data` block. It's been a lot.

This is a natural break point for this series. In the not-too-distant future I plan on extending this series to explore simple multilevel models for **brms** and Stan, where we'll discuss the centered and non-centered parameterizations, and go wild with parameter vectors. Until then, happy coding, friends!

## Thank the reviewers

I'd like to publicly acknowledge and thank

* [Riccardo Fusaroli](https://pure.au.dk/portal/en/persons/fusaroli%40cc.au.dk) and
* [Stephen J. Wild](https://sjwild.github.io/)

for their kind efforts reviewing the draft of this post. Go team!

Do note the final editorial decisions were my own, and I do not think it would be reasonable to assume my reviewers have given blanket endorsements of the current version of this post.

## Session information

```{r}
sessionInfo()
```

```{r, fig.width = 6, fig.height = 4, echo = FALSE, eval = FALSE, warning = FALSE}
xmax <- max(d$height) |> ceiling()
m_x <- mean(d$height)
m_y <- mean(d$weight)
ymax <- max(d$weight)
ymin <- as_draws_df(brm3) |> summarise(quantile(b_Intercept, 0.025)) |> pull()

p1 <- data.frame(height = 0:xmax) |> 
  add_epred_draws(brm3) |> 
  rename(weight = .epred) |> 
  
  ggplot(aes(x = height, y = weight)) +
  stat_lineribbon(.width = 0.95,
                  color = "steelblue4", fill = "steelblue1", linewidth = 1/2) +
  geom_point(data = d,
             size = 3/4) +
  geom_hline(yintercept = m_y, color = "white") +
  geom_vline(xintercept = m_x, color = "white") +
  theme(panel.grid = element_blank())

# p2 <- as_draws_df(brm3) |> 
#   ggplot(aes(y = Intercept)) +
#   geom_hline(yintercept = m_y, color = "white") +
#   stat_halfeye(.width = 0.95, 
#                color = "steelblue4", fill = "steelblue1", size = 1/10) +
#   scale_x_continuous("Intercept", breaks = NULL, expand = expansion(mult = 0.15), position = "top") +
#   scale_y_continuous(NULL, breaks = NULL, limits = c(ymin, ymax))

p2 <- as_draws_df(brm3) |> 
  pivot_longer(cols = contains("Intercept")) |> 
  
  ggplot(aes(y = value, x = name)) +
  geom_hline(yintercept = m_y, color = "white") +
  stat_pointinterval(.width = 0.95, 
               color = "steelblue4", size = 1/10) +
  scale_x_discrete(NULL, position = "top") +
  scale_y_continuous(NULL, breaks = NULL, limits = c(ymin, ymax)) +
  theme(axis.text.x = element_text(angle = 270, hjust = 1),
        panel.grid = element_blank())

p1 + p2 +
  plot_layout(widths = c(12, 1))

# ggsave("height_cross-featured.jpg", width = 3.6, height = 3.5, units = "in")
```

## References 

